Draft -- Discriminating Seagrasses From Green Macroalgae in European Intertidal areas using high resolution multispectral drone imagery -- Draft

### 2024-10-11 docs/_tex/index.tex
add updates
pierre comments

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\usepackage{bookmark}

@@

\begin{abstract}
Coastal areas support seagrass meadows, which offer crucial ecosystem
services including erosion control and carbon sequestration. However,
these areas are increasingly impacted by human activities, leading to
habitat fragmentation and seagrass decline. In situ surveys,
traditionally performed to monitor these ecosystems face limitations on
temporal and spatial coverage, particularly in intertidal zones,
prompting the addition of satellite data within monitoring programs.
Yet, satellite remote sensing struggles with spatial and spectral
resolution, making it difficult to discriminate seagrass from other
macrophytes in highly heterogenous meadows. Drone (unmanned aerial
vehicles -- UAV) images at a very high spatial resolution offer a
promising solution to address challenges related to spatial
heterogeneity and intrapixel mixture. This study focuses on using drone
acquisitions with a ten spectral band sensor mirroring those of
Sentinel-2, for mapping intertidal macrophytes and effectively
discriminating between seagrass and green macroalgae. Nine drone flights
were conducted at two different altitudes (12 m and 120 m) across
heterogeneous intertidal European habitats in France and Portugal. Low
altitude flights were used to train a Deep Learning classifier based on
Neural Networks to discrimintate among five taxonomic classes of
intertidal vegetation: Magnoliopsida (Seagrass), Chlorophyceae (Green
macroalgae), Phaeophyceae (Brown algae), Rhodophyceae (Red macroalgae)
and benthic Bacillariophyceae (Diatoms). Classification of drone imagery
resulted in an overall accuracy of 94\% across all the sites and images,
covering a total area of 467 000 m². The model exhibited an accuracy of
96.4\% in identifying seagrass. Importantly, seagrass and green algae
can be discriminated, although they share the same pigment composition.
As the algorithm was developed for a multispectral camera with ten
spectral bands in the visible and near-infrared, it could be adapted to
the Multi-Spectral Instrument (MSI) onboard Sentinel-2 thus offering
promising perspectives for satellite remote sensing of intertidal
biodiversity over lager scales.
\end{abstract}

@@

Coastal areas are vital hotspots for marine biodiversity, with
intertidal seagrass meadows playing a crucial role at the interface
between the land and oceans \citep{unsworth2022}. Seagrass meadows
provide a myriad of ecosystem services, including carbon sequestration,
oxygen production, protection against sea-level rise and coastline
erosion, and mitigation of eutrophication
\citep{unsworth2022, sousa2019blue}. They serve as vital habitats for a
diverse array of marine and terrestrial species, providing living,
breeding, and feeding grounds
\citep{gardner2018, Zoffoli2022, jankowska2019}. Due to the
concentration of human activities in coastal zones, seagrass meadows are
directly exposed to and impacted by anthropogenic pressures. Global
regression and fragmentation of seagrass meadows are currently observed
due to climate change, diseases, urbanization, land reclamation,
dredging, competition with alien species, and reduction in water quality
\citep{nguyen2021, soissons2018, orth2006, lin2018, duffy2019, rasheed2011long, chefaoui2018dramatic, sousa2019blue}.
Both habitat fragmentation and reduction, in turn, can severely
compromise the effectiveness of ecosystem services provided by seagrass
meadows. While improvements in water quality and hydrodynamics have been
recently reported in Europe, allowing an overall recovery of seagrass
ecosystems at local and European scales, many coastal waters worldwide
are still subjected to strong eutrophication processes
\citep{deSantos2019, Zoffoli2021, sousa2019blue}. Coastal eutrophication
has been associated to excessive accumulation of green macroalgae,
so-called green tides \citep{devlin2023nutrients}. Green tides produce
shade and suffocation over seagrass individuals, thus threatening the
health of seagrass ecosystems \citep{wang2022}.

The importance of seagrass meadows and the variety of ecosystem services
they provide have led to the enhancement of both global and regional
programs to monitor Essential Oceanic Variable (EOVs) such as seagrass
composition \citep{Miloslavich2018}, as well as Essential Biodiversity
Variable (EBVs) such as seagrass taxonomic diversity, species
distribution, population abundance, and phenology \citep{Pereira2013}.
Traditionally, indicators of seagrass status have been quantified using
\emph{in situ} measurements. However, the acquisition of these
measurements in intertidal zones is notoriously challenging. Intertidal
seagrass meadows are only exposed during low tide and can be situated in
difficult-to-reach mudflats, potentially leading to inaccurate and
limited estimations with conventional sampling techniques
\citep{nijland2019}. Satellite observations have been proven effective
in complementing \emph{in situ} sampling, allowing for the near
real-time and consistent retrieval of seagrass EOVs and EBVs over
extensive meadows. \citep{Zoffoli2021, xu2021, Traganos2018, coffer2023}

While satellite remote sensing (RS) provides temporally consistent
observations over large spatial scales, its utilization over intertidal
areas is limited by several constraints. Satellite missions with a high
temporal resolution (e.g.~daily MODIS observation) are limited by a too
coarse spatial resolution (\textgreater100 m) to accurately map patchy
seagrass meadows. Missions with a high spatial resolution such as
Sentinel-2 (10 m) or Landsat8/9 (30 m) can be limited by low spectral
resolution. The limited number of spectral bands challenges accurate
discrimination of seagrass from other co-existing macrophytes. In
particular Chlorophyceae (green algae) and marine Magnoliopsida
(seagrass) share the same pigment composition
\citep{ralph2002, Douay2022}. Therefore, to someone not specialized in
the field, their spectral signatures may appear to be alike
\citep{Davies2023, bannari2022}. Recently, using advanced
machine-learning algorithms trained with a large hyperspectral library
of more than 300 field reflectance spectra, \citep{Davies2023}
demonstrated that it was possible to discriminate Magnoliopsida from
Chlorophyceae using radiometric data acquired at Sentinel-2 's spectral
resolution. However the application of this approach to satellite RS
remains to be validated. Moreover patches of green algae can develop at
small spatial scales that are not observable using non-commercial
satellite imagery \citep{tuya2013}, especially during the initial stage
of a green tide.

Drones (Unmanned Aerial Vehicles -- UAVs) can potentially fill the data
gaps left by satellite RS and \emph{in situ} measurements, due to their
ability to provide spatially-explicit observations at very high spatial
resolutions (pixel size from mm to cm) while capturing data at
multi-spectral resolution \citep{fairley2022drone, oh2017use}. The
versatility of drones allows for their application across a diverse
thematic range , from coastal zone management
\citep{adade2021, casella2020, angnuureng2022} to mapping species
distribution
\citep{joyce2023, tallam2023, Roca2022, Roman2021, Brunier2022Topographic, sousa2019blue}.
However, when applied to coastal habitat mapping, most case studies were
previously limited to a single flight, restricting the generalizability
of their application over wider geographical scales
\citep{Roman2021, collin2019improving, rossiter2020uav, Brunier2022Topographic}.
The present study aimed at analyzing the potential of multispectral
drone RS to map intertidal macrophytes over a diverse biogeographical
range, with a particular focus on discriminating Magnoliopsida and
Chlorophyceae (Seagrass and Green Algae, respectively). Ten drone
flights were performed over soft-bottom intertidal areas along the
Atlantic coastlines of two European countries (France and Portugal),
covering a wide range of habitats, from monospecific seagrass meadows to
meadows mixed with green, or red macroalgae. A deep learning algorithm
was trained and validated for macrophyte discrimination, emphasizing
applicability across diverse sites without a loss of prediction
accuracy. In addition, the results obtained from drone-based
multispectral remote sensing will be spatially degraded to satellite
resolutions, and the effect of this resolution on the model's
predictions will be investigated, providing insights for coastal habitat
mapping using satellites.

@@

At each location, a DJI Matrice 200 quadcopter drone equipped with a
Micasense RedEdge Dual MX multispectral camera was flown to take 1.2
million pixel reflectance photographs with ten spectral bands ranging
from the blue to the near infrared (NIR): 444, 475, 531, 560, 650, 668,
705, 717, 740 and 840 nm. To ensure consistent lighting conditions
across flight paths, the drone's trajectory was aligned to maintain a
solar azimuth angle of 90 degrees. An overlap of 70\% and 80\% (side and
front respectively) between each image was set for each flight. A
downwelling light sensor (DLS2) was used to acquire irradiance data
concomitantly with the camera measurements. Raw data were calibrated in
reflectance using a calibration panel reflective at \textasciitilde50\%
provided by the manufacturer. Across all sites, flights were made at two
different altitudes : 12 m or/and 120 m. (Table~\ref{tbl-flights}).

@@

A structure-from-motion photogrammetry software \citep[Agisoft
Metashape,][]{agisoft}was used to process images to obtain multispectral
orthomosaics of each flight. The process for orthomosaicking was
identical for every flight. First, tying key points were detected inside
of each image and between overlapping images in order to obtain a sparse
point cloud. This cloud was cleaned using reprojection accuracy metric
in order to remove noisy points. A dense point cloud was then produced
using a structure from motion algorithm. A surface interpolation of this
dense point cloud was made to obtain a digital surface model (DSM), used
to reconstruct the multispectral ortho-image \citep{nebel2020review}.
Low altitude drone flights produced ortho-images with a very high
spatial resolution (8 mm per pixel), making it efficient to visually
distinguish between the various types of vegetation. High altitude
flights on the other hand allowed to cover large areas and produced
images with a pixel size of 80 mm (Table~\ref{tbl-flights}).

@@

The workflow of this study revolves around two distinct flight heights
(12 m and 120 m, Figure~\ref{fig-workflow}) where ensuring consistency
between reflectances at both heights is crucial. This comparison was
conducted at sites where low and high altitude flights overlapped. The
low altitude flights were resampled to the same spatial resolution and
grid as the high flights using a median resampling method. Reflectance
values were then extracted, and a scatterplot was generated.
Subsequently, the slope of the linear model and the coefficient of
determination (R²) were computed.

@@

\subsection{Impact of the spatial resolution in the
prediction}\label{impact-of-the-spatial-resolution-in-the-prediction}

To evaluate the impact of spatial resolution on the model's output, we
resampled the reflectance of each flight's orthomosaic using the average
method from the Terra package in R. DISCOV was then applied to these
resampled rasters, and the results were compared to the original model
predictions. For each resolution and vegetation class, we calculated
predicted area loss, where a score of 0 indicates no area loss during
resampling, and a score of 100 indicates complete loss of vegetation.

We used a Generalized Linear Model (GLM) with a Gaussian family and
identity link function to examine the relationship between pixel
resolution, vegetation class, and their interaction on the response
variable (the loss of vegetation area). The model formula was as
follows:

\begin{equation}\phantomsection\label{eq-gam}{
%loss \sim  Resolution \times VegetationClass
}\end{equation}

@@

The key aspect of the workflow adopted in the present study is the
mapping at two different altitudes (12 m and 120 m), resulting in two
distinct resolutions for the same area (8 mm and 80 mm). The
high-resolution flight was used to estimate the sub-pixel composition
for each pixel of the lower-resolution flight. Consequently, within each
pixel of the high-altitude flights, the contribution of each vegetation
class (\% cover) was obtained, and a kernel density plot was generated.
This plot provided a visual representation of the model's behavior in
mixed vegetation scenarios. It helps to understand the minimum
vegetation cover of a given class within a pixel necessary for the model
to confidently predict that class.

@@

In this study, drone flights were conducted at two different altitudes
(12 m and 120 m) to construct the neural network model. At the sites
where the flights at both altitudes overlapped, reflectance was
compared. Overall there was a good agreement between the two altitudes
(RMSE: 0.027 ; Figure~\ref{fig-CompareRef}).

@@

There was a slight underestimation for raw reflectance values in the
high-altitude flight, particularly for higher reflectance values
(Figure~\ref{fig-CompareRef} A). Since both flights were conducted over
vegetation areas, the highest reflectance values correspond to the
infrared part of the spectrum. This difference was not present when
reflectance values have been standardized (Equation~\ref{eq-std} ;
Figure~\ref{fig-CompareRef} B).

@@

\subsection{Effect of the spatial resolution on the
prediction}\label{effect-of-the-spatial-resolution-on-the-prediction}

Figure~\ref{fig-pixelsize} illustrates the relationship between
predicted area loss for various vegetation types and map resolution. At
a fine resolution of 1m, changes in the retrieved area for each
vegetation type are minimal. Green algae shows the highest loss rate,
with 1.2\% area lost compared to the raw resolution (8cm). As the
resolution coarsens to 10m, area loss becomes significant, with green
algae again experiencing the greatest reduction (12\% compared to 8cm)
and seagrass showing the smallest loss (1.3\%). At a resolution of 30m,
all green algae have been lost (100\% compared to 8cm), while seagrass
experiences a relatively small reduction of 11\%. Brown algae and red
algae show moderate declines, with losses at 30m resolution reaching
approximately 37\% and 59\%, respectively.

@@

A percent cover of at least 80\% was sufficient to have all the pixels
of high altitude flights correctly classified, with the exception of
Magnoliopsida that required a higher cover (\textgreater90 \%) to be
well classified. Concerning the probability of each class, there is a
linear relationship between the percent cover and the confidence of the
model to predict the class. To predict Chlorophyceae with a model
likelihood of 0.85, a cover of 93 \% was needed, 90 \% for
Magnoliopsida, 92 \% for Rhodophyceae and 97 \% for Bacillariophyceae.
When the vegetation cover of a given class was 100 \%, coarser
high-flight pixels were well classified for all the classes except for
Bare Sediment, which was only well classified 80\% of the time. This
phenomenon may be attributed to the time gap between the two flights,
allowing for microphytobenthos migration to the surface during low tide,
consequently altering the model's classification from bare sediment to
Bacillariophyceae.

@@

The primary objective of this study was to develop a method for the
accurate classification of macrophytes at low tide on tidal flats,
specifically focusing on distinguishing between Chlorophyceae (green
macroalgae) and marine Magnoliopsida (seagrasses) using a multispectral
resolution. The discrimination of seagrasses from green macroalgae
presents is spectrally challenging
\citep{oiry2021using, bannari2022, veettil2020opportunities}. These two
macrophytes share a similar pigment composition: chlorophyll-a (common
to all vegetation types), chlorophyll-b (an additional photosynthetic
pigment), and accessory carotenoids such as zeaxanthin, lutein and
neoxanthin (Figure~\ref{fig-Pigm}). Their spectral responses could be
close, particularly at a multispectral resolution. Seagrass and green
macroalgae frequently co-occur in intertidal areas, and can intermingle
within a remote sensing pixel if the spatial resolution is too low.
Here, the issue of intra-pixel mixing was resolved thanks to the very
high spatial resolution of the drone. In this study the risk of spectral
confusion was avoided with a machine-learning approach exploiting a
neural networks classifier. Our drone flights and a recent study based
on \emph{in situ} radiometry, suggested that a sensor with at least
eight spectral bands ranging from 500 to 850 nm and including a green
band at 530 nm and a RedEdge band at 730 nm was crucial to accurately
discriminate green macroalgae from seagrasses \citep{Davies2023}.

@@

This resolution-area relationship has important implications for
satellite platforms like Sentinel-2 and Landsat, which are commonly used
in marine and coastal vegetation studies. Both satellites offer
medium-resolution imagery, with pixel sizes of 10m and 30m,
respectively. While these resolutions are suitable for broad-scale
environmental monitoring, they may struggle to capture finer-scale
heterogeneity, particularly in fragmented vegetation like green algae.
The findings suggest that, while coarser resolutions like those of
Landsat may be adequate for more homogeneous types, such as seagrass,
higher resolutions are essential for accurately mapping patchier species
like green algae. These findings have direct implications for
environmental management and conservation planning. Overlooking
fine-scale vegetation features, such as those seen in green algae, could
result in inadequate protection or restoration efforts, particularly in
ecologically sensitive coastal zones, as early stages of green tides
could be challenging to monitor at coarse resolutions.

Higher-resolution imagery offers more accurate vegetation mapping but
comes with trade-offs. As resolution increases, data costs rise, and
processing becomes more resource-intensive due to the larger file sizes
and computational demands. Additionally, high-resolution data requires
more storage and can slow down real-time applications. For large-scale
monitoring or homogeneous vegetation types, medium-resolution platforms
like Sentinel-2 are often sufficient. However, when mapping fragmented
vegetation like green algae, the precision provided by higher-resolution
imagery is crucial, despite the additional costs and processing
challenges it imposes.

>>>>

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\usepackage{bookmark}

@@

\begin{abstract}
Coastal areas support seagrass meadows, which offer crucial ecosystem
services including erosion control and carbon sequestration. However,
these areas are increasingly impacted by human activities, leading to
habitat fragmentation and seagrass decline. In situ surveys,
traditionally performed to monitor these ecosystems face limitations on
temporal and spatial coverage, particularly in intertidal zones,
prompting the addition of satellite data within monitoring programs.
Yet, satellite remote sensing can be limited by too coarse spatial
and/or spectral resolution, making it difficult to discriminate seagrass
from other macrophytes in highly heterogenous meadows. Drone (unmanned
aerial vehicles -- UAV) images at a very high spatial resolution offer a
promising solution to address challenges related to spatial
heterogeneity and intrapixel mixture. This study focuses on using drone
acquisitions with a ten spectral band sensor similar to that onboard
Sentinel-2, for mapping intertidal macrophytes at low tide (i.e.~during
a period of emersion) and effectively discriminating between seagrass
and green macroalgae. Nine drone flights were conducted at two different
altitudes (12 m and 120 m) across heterogeneous intertidal European
habitats in France and Portugal, providing multispectral reflectance
observation at very high spatial resolution (respectively at 8 mm and 80
mm). Taking advantage of their extremly high spatial resolution; the low
altitude flights were used to train a Deep Learning classifier based on
Neural Networks to discrimintate among five taxonomic classes of
intertidal vegetation: Magnoliopsida (Seagrass), Chlorophyceae (Green
macroalgae), Phaeophyceae (Brown algae), Rhodophyceae (Red macroalgae)
and benthic Bacillariophyceae (Diatoms), and validated using concomitant
field measurements. Classification of drone imagery resulted in an
overall accuracy of 94\% across all sites and images, covering a total
area of 467 000 m². The model exhibited an accuracy of 96.4\% in
identifying seagrass. In particular , seagrass and green algae can be
discriminated. The very high spatial resolution of the drone data made
ot possoble to assess the influence of spatial resolution on the
classification outputs, showing a limited loss in seagrass detection up
to about 10m. Altogether, out findings suggest that the Multi-Spectral
Instrument (MSI) onboard Sentinel-2 offers a relevant trade-off between
its spatial and spectral resolution, thus offering promising
perspectives for satellite remote sensing of intertidal biodiversity
over lager scales.
\end{abstract}

@@

Coastal areas are vital hotspots for marine biodiversity, with
intertidal seagrass meadows playing a crucial role at the interface
between land and ocean \citep{unsworth2022}. Seagrass meadows provide a
myriad of ecosystem services, including carbon sequestration, oxygen
production, protection against sea-level rise and coastline erosion, and
mitigation of eutrophication \citep{unsworth2022, sousa2019blue}. They
serve as vital habitats for a diverse array of marine and terrestrial
species, providing living, breeding, and feeding grounds
\citep{gardner2018, Zoffoli2022, jankowska2019}. Due to the
concentration of human activities in coastal zones, seagrass meadows are
directly exposed to and impacted by anthropogenic pressures. Global
regression and fragmentation of seagrass meadows are currently observed
due to climate change, diseases, urbanization, land reclamation,
dredging, competition with alien species, and reduction in water quality
\citep{nguyen2021, soissons2018, orth2006, lin2018, duffy2019, rasheed2011long, chefaoui2018dramatic, sousa2019blue}.
Both habitat fragmentation and reduction, in turn, can severely
compromise the effectiveness of ecosystem services provided by seagrass
meadows. While improvements in water quality and hydrodynamics have been
recently reported in Europe, allowing an overall recovery of seagrass
ecosystems at local and European scales, many coastal waters worldwide
are still subjected to strong eutrophication processes
\citep{deSantos2019, Zoffoli2021, sousa2019blue}. Coastal eutrophication
has been associated to excessive accumulation of green macroalgae,
so-called green tides \citep{devlin2023nutrients}. Green tides produce
shade and suffocation over seagrass individuals, thus threatening the
health of seagrass ecosystems \citep{wang2022}.

The importance of seagrass meadows and the variety of ecosystem services
they provide have led to the enhancement of both global and regional
programs to monitor Essential Oceanic Variable (EOVs) such as seagrass
composition \citep{Miloslavich2018}, as well as Essential Biodiversity
Variable (EBVs) such as seagrass taxonomic diversity, species
distribution, population abundance, and phenology \citep{Pereira2013}.
Traditionally, indicators of seagrass status have been quantified using
\emph{in situ} measurements. However, the acquisition of field
measurements in intertidal zones is notoriously challenging. Intertidal
seagrass meadows are only exposed during low tide and can be situated in
difficult-to-reach mudflats, potentially leading to inaccurate and
limited estimations with conventional sampling techniques
\citep{nijland2019}. Satellite observations have been proven effective
in complementing \emph{in situ} sampling, allowing for near real-time
and consistent retrieval of seagrass EOVs and EBVs over extensive
meadows \citep{Zoffoli2021, xu2021, Traganos2018, coffer2023}.

While satellite remote sensing (RS) provides temporally consistent
observations over large spatial scales, its utilization over intertidal
areas is limited by several constraints. Satellite missions with a high
temporal resolution (e.g.~daily MODIS observation) are limited by a too
coarse spatial resolution (\textgreater100 m) to accurately map patchy
seagrass meadows. Missions with a high spatial resolution such as
Sentinel-2 (10 m) or Landsat8/9 (30 m) can be limited by low spectral
resolution. The limited number of spectral bands challenges accurate
discrimination of seagrass from other co-existing macrophytes. In
particular, Chlorophyceae (green algae) and marine Magnoliopsida
(seagrass) share the same pigment composition
\citep{ralph2002, Douay2022}, resulting in a similar spectral signature
in therms of reflectance, especially in the visible range
\citep{Davies2023, bannari2022}. Recently, using advanced
machine-learning algorithms trained with a large hyperspectral library
of more than 300 field reflectance spectra, \citep{Davies2023}
demonstrated that it was possible to discriminate Magnoliopsida from
Chlorophyceae using reflectance spectra at Sentinel-2 's spectral
resolution. However the application of this approach to satellite RS
remains to be validated. Moreover patches of green algae can develop at
small spatial scales that are not observable using Sentinel-2 and/or
Landsat-8/9 images \citep{tuya2013}, especially during the initial stage
of a green tide.

Drones (Unmanned Aerial Vehicles -- UAVs) can potentially fill the data
gaps left by satellite RS and \emph{in situ} measurements, due to their
ability to provide spatially-explicit observations at very high spatial
resolutions (pixel size from mm to cm) while capturing data at
multi-spectral resolution \citep{fairley2022drone, oh2017use}. The
versatility of drones allows for their application across a diverse
thematic range , from coastal zone management
\citep{adade2021, casella2020, angnuureng2022} to mapping species
distribution
\citep{joyce2023, tallam2023, Roca2022, Roman2021, Brunier2022Topographic, sousa2019blue}.
However, when applied to coastal habitat mapping, previous case studies
were mostly limited to a limited number of flights over a single study
site, restricting the generalizability of their application over wider
geographical scales
\citep{Roman2021, collin2019improving, rossiter2020uav, Brunier2022Topographic}.
The present study aimed at analyzing the potential of multispectral
drone RS to map intertidal macrophytes over a diverse biogeographical
range, with a particular focus on discriminating Magnoliopsida and
Chlorophyceae (Seagrass and Green Algae, respectively). Nine drone
flights were performed over soft-bottom intertidal areas along the
Atlantic coastlines of two European countries (France and Portugal),
covering a wide range of habitats, from monospecific seagrass meadows to
meadows mixed with green, or red macroalgae. A deep learning algorithm
was trained and validated for macrophyte discrimination, emphasizing
applicability across diverse sites without a loss of prediction
accuracy. The classification maps obtained at a very high spatial
resolution with the drone were spatially degraded to satellite
resolutions, making it possible to asses the effect of spatial
resolution on classification accuracy; and provide insights for coastal
habitat mapping using satellite remote sensing.

@@

At each location, a DJI Matrice 200 quadcopter drone equipped with a
Micasense RedEdge Dual MX multispectral camera was flown to take 1.2
million pixel reflectance photographs with ten spectral bands ranging
from the blue to the near infrared (NIR): 444, 475, 531, 560, 650, 668,
705, 717, 740 and 840 nm. To ensure consistent lighting conditions
across flight paths, the drone's trajectory was aligned to maintain a
solar azimuth angle of 90 degrees. An overlap of 70\% and 80\% (side and
front respectively) between each image was set for each flight. A
downwelling light sensor (DLS2) was used to acquire irradiance data
concomitantly with the camera measurements. Raw data were calibrated in
reflectance using a calibration panel reflective at \textasciitilde50\%
provided by the manufacturer. Across all sites, flights were made at two
different altitudes : 12 m or/and 120 m, with a spatial resolution of 8
mm and 80 mm, respectively (Table~\ref{tbl-flights}).

@@

A structure-from-motion photogrammetry software \citep[Agisoft
Metashape,][]{agisoft} was used to process images to obtain
multispectral orthomosaics of each flight. The process for
orthomosaicking was identical for every flight. First, tying key points
were detected inside of each image and between overlapping images in
order to obtain a sparse point cloud. This cloud was cleaned using
reprojection accuracy metric in order to remove noisy points. A dense
point cloud was then produced using a structure from motion algorithm. A
surface interpolation of this dense point cloud was made to obtain a
digital surface model (DSM), used to reconstruct the multispectral
ortho-image \citep{nebel2020review}. Low altitude drone flights produced
ortho-images with a very high spatial resolution (8 mm per pixel),
making it efficient to visually distinguish between the various types of
vegetation. High altitude flights allowed to cover larger areas and
produced images with a pixel size of 80 mm (Table~\ref{tbl-flights}).

@@

The workflow of this study revolves around two distinct flight heights
(12 and 120 m, Figure~\ref{fig-workflow}) where ensuring consistency
between reflectances at both heights is crucial. This comparison was
conducted at sites where low and high altitude flights overlapped. The
low altitude flights were resampled to the same spatial resolution and
grid as the high flights using a median resampling method. Reflectance
values were then extracted, and a scatterplot was generated.
Subsequently, the slope of the linear model and the coefficient of
determination (R²) were computed.

@@

\subsection{Influence of the spatial resolution on
classification}\label{influence-of-the-spatial-resolution-on-classification}

To evaluate the influence of spatial resolution on the model's output,
we resampled the drone orthomosaics from their native resolution (8 cm
for the high altitude flights) using the ``average'' method from the
Terra package in R. DISCOV was then applied to these resampled rasters,
and the results were compared to the original model predictions. For
each resolution and vegetation class, we calculated the predicted area
loss, where a score of 0 indicates no area loss during spatial
resampling, and a score of 100 indicates complete loss of the vegetation
class.

We used a Generalized Linear Model (GLM) with a Gaussian family and
identity link function to examine the relationship between pixel
resolution, vegetation class, and their interaction on the response
variable (the loss of vegetation area within a given class). The model
formula was as follows:

\begin{equation}\phantomsection\label{eq-gam}{
VegetationLoss \sim  Resolution \times VegetationClass
}\end{equation}

@@

The key aspect of the workflow adopted in the present study is the
mapping at two different altitudes (12 and 120 m), resulting in two
distinct resolutions for the same area (8 and 80 mm; respectively). The
high-resolution flight was used to estimate the sub-pixel composition
for each pixel of the lower-resolution flight. Consequently, within each
pixel of the high-altitude flights, the contribution of each vegetation
class (\% cover) was obtained, and a kernel density plot was generated.
This plot provided a visual representation of the model's behavior in
mixed vegetation scenarios. It helps to understand the minimum
vegetation cover of a given class within a pixel necessary for the model
to confidently predict that class.

@@

In this study, drone flights were conducted at two different altitudes
(12 and 120 m) to construct the neural network model. At the sites where
the flights at both altitudes overlapped, reflectance was compared.
Overall there was a good agreement between the two altitudes (RMSE:
0.027 ; Figure~\ref{fig-CompareRef}).

@@

There was a slight underestimation for raw reflectance values in the
high-altitude flight, particularly for higher reflectance values
(Figure~\ref{fig-CompareRef} A). Since both flights were conducted over
vegetated areas, the highest reflectance values correspond to the
infrared part of the spectrum. This difference was not present when the
reflectance has been standardized (Equation~\ref{eq-std} ;
Figure~\ref{fig-CompareRef} B).

@@

\subsection{Effect of spatial resolution on the
classification}\label{effect-of-spatial-resolution-on-the-classification}

Figure~\ref{fig-pixelsize} illustrates the relationship between
predicted area loss for various vegetation types and map resolution. At
a fine resolution of 1m, changes in the retrieved area for each
vegetation type are minimal. Green algae shows the highest loss rate,
with 1.2\% area lost compared to the native resolution (80 mm). As the
resolution coarsens to 10m, area loss becomes significant, with green
algae again experiencing the greatest reduction (12\% compared to 8cm)
and seagrass showing the smallest loss (1.3\%). At a resolution of 30m,
all green algae have been lost (100\% compared to 8cm), while seagrass
experiences a relatively small reduction of 11\%. Brown algae and red
algae show moderate declines, with losses at 30m resolution reaching
approximately 37\% and 59\%, respectively.

@@

A percent cover of at least 80\% was sufficient to have all the 80 mm
pixels correctly classified, with the exception of Magnoliopsida that
required a higher cover (\textgreater90 \%) to be accurately classified.
Concerning the probability of each class, there is a linear relationship
between the percent cover and the confidence of the model to predict the
class. To predict Chlorophyceae with a model likelihood of 0.85, a cover
of 93 \% was needed, 90 \% for Magnoliopsida, 92 \% for Rhodophyceae and
97 \% for Bacillariophyceae. When the vegetation cover of a given class
was 100 \%, coarser high-flight pixels were correctly classified for all
the classes except for Bare Sediment, which was only correctly
classified 80\% of the time. This phenomenon may be attributed to the
time gap between the two flights, allowing for microphytobenthos
migration to the surface during low tide, consequently altering the
model's classification from bare sediment to Bacillariophyceae.

@@

The primary objective of this study was to develop a method for the
accurate classification of emerged macrophytes observed during low tide
on tidal flats, specifically focusing on distinguishing between
Chlorophyceae (green macroalgae) and marine Magnoliopsida (seagrasses)
using a multispectral resolution. The discrimination between seagrasses
and green macroalgae is challenging due to their optical similarity in
the visible range
\citep{oiry2021using, bannari2022, veettil2020opportunities}. These two
macrophytes share a similar pigment composition: chlorophyll-a (common
to all vegetation types), chlorophyll-b (an additional photosynthetic
pigment), and accessory carotenoids such as zeaxanthin, lutein and
neoxanthin (Figure~\ref{fig-Pigm}). Their spectral responses could be
close, particularly at a multispectral resolution. Seagrass and green
macroalgae frequently co-occur in intertidal areas, and can intermingle
within a remote sensing pixel if the spatial resolution is too low.
Here, the issue of intra-pixel mixing was resolved thanks to the very
high spatial resolution of the drone (from 8 to 80 mm). In this study
the risk of spectral confusion was avoided with a machine-learning
approach exploiting a neural networks classifier. Our drone flights and
a recent study based on \emph{in situ} radiometry, suggested that a
sensor with at least eight spectral bands ranging from 500 to 850 nm,
and including a green band at 530 nm and a RedEdge band at 730 nm, was
crucial to accurately discriminate green macroalgae from seagrasses
\citep{Davies2023}.

@@

This resolution-area relationship has important implications for
satellite missions like Sentinel-2 and Landsat, which are commonly used
in marine and coastal vegetation studies. Both satellites offer
high-resolution imagery, with pixel sizes of 10m and 30m, respectively.
While these resolutions are suitable for broad-scale environmental
monitoring, they may be too coarse to capture finer-scale heterogeneity,
particularly in fragmented vegetation like green algae. Our findings
suggest that, while the 30m resolution of Landsat may be adequate for
homogeneous vegetation types, such as seagrass, a higher resolution is
essential for accurately mapping patchy vegetation like green algae.
These findings have direct implications for environmental management and
conservation planning. Overlooking fine-scale vegetation features, such
as those seen in green algae, could result in inadequate protection or
restoration efforts, particularly in ecologically sensitive coastal
zones, as early stages of green tides could be challenging to monitor at
coarse resolutions.

Very high-resolution imagery offers more accurate vegetation mapping but
comes with trade-offs. As resolution increases, data costs rise, and
processing becomes more resource-intensive due to the larger file sizes
and computational demands. Additionally, high-resolution data requires
more storage and can slow down real-time applications. For large-scale
monitoring or homogeneous vegetation types, 10 m resolution of S2/MSI or
even the 30 m of Landsat/OLI are often sufficient. However, when mapping
fragmented vegetation like green algae, the precision provided by
higher-resolution imagery is crucial, despite the additional costs and
processing challenges it imposes.
